---
title: "Project 2"
author: "Group Members' Names"
date: "Date"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
require("knitr")
datadir <- "C:/Users/JK/Google Drive/Acedemics/College/Fourth Year/Semester 7/Linear Statistical Models/Data/AirQualityData"
sourcedir <-"C:/Users/JK/Google Drive/Acedemics/College/Fourth Year/Semester 7/Linear Statistical Models/Code"
opts_knit$set(root.dir = sourcedir)
library(forecast)
library(mtsdi)
library(MTS)
```

```{r setup, include = FALSE}
# Camryn Set Up
require("knitr")
datadir <- "/Users/camrynburley/Desktop/UVA/SYS4021/Data/AirQuality"
sourcedir <- "/Users/camrynburley/Desktop/UVA/SYS4021/RCode"
opts_knit$set(root.dir = sourcedir)
library(forecast)
library(mtsdi)
library(MTS)
```

# Load Data and Impute Missing Values
```{r, warning=FALSE}
setwd(datadir)

airquality = read.csv('AirQualityUCI.csv')

# replace -200 with NA
airquality[airquality == -200] <- NA

# convert integer type to numeric
intcols = c(4,5,7,8,9,10,11,12)
for(i in 1:length(intcols)){
  airquality[,intcols[i]] <- as.numeric(airquality[,intcols[i]])
}

setwd(sourcedir)

# create new data frame with just CO and NO2
AQdata = airquality[,c(3,10)]

# impute missing air quality data
f <- ~ CO.GT. + NO2.GT.
t <- c(seq(1,dim(AQdata)[1],1))
i <- mnimput(f, AQdata, eps=1e-3, ts=TRUE, method='gam', 
             ga.control=list(formula=paste(names(AQdata)[c(1:3)],'~ns(t,2)')))

# set airquality to imputed data
AQdata <- i$filled.dataset

# aggregate to daily maxima for model building
dailyAQ <- aggregate(AQdata, by=list(as.Date(airquality[,1],"%m/%d/%Y")), FUN=max)
```

# Create Testing and Training Data
```{r}
# the index to split the data on
separationIndex <- nrow(dailyAQ)-6

# split the data
air.train <- dailyAQ[1:separationIndex-1,]
air.test <- dailyAQ[separationIndex:nrow(dailyAQ),]

# making sure no data was duplicated or lost
a <- nrow(dailyAQ)
b <- nrow(air.train)
c <- nrow(air.test)

a
b
c
```

# Data Summary
```{r}
summary(air.train)
```
The data is from March 2004 through April 2005. We are modeling the ambient daily maximum carbon monoxide (CO) and nitrogen dioxide (NO2) concentrations. 

# Univariate Time Series for CO
## Visualize the Data
```{r}
# create time series
CO.ts <- ts(air.train$CO.GT.)

# visualize raw data
plot(CO.ts)

# periodogram
# this pumps out a different graph sometimes
  # is this because it is doing fast Fourier Transfer??
pg.CO <- spec.pgram(CO.ts,spans=9,demean=T,log='no')

# acf and pacf
acf(CO.ts)
pacf(CO.ts)
```

based on the periodogram there is not one main season, but possible multiple
-if there is seasonality it will be based on a complex wave

No obvious trend or seasonality

## Seasonality
### Finding Potential Periods
```{r}
# sort the frequencies based on influence
sorted.spec <- sort(pg.CO$spec, decreasing=T, index.return=T)

# convert to periods
sorted.omegas <- pg.CO$freq[sorted.spec$ix]
sorted.Ts <- 1/pg.CO$freq[sorted.spec$ix]

# the cutoff for influencial
CO.pg.cutoff <- 10

# the top periods
print('top periods')
sorted.Ts[1:CO.pg.cutoff]

# top frequencies
## to double check that this makes sense based on periodogram
print('top frequencies')
sorted.omegas[1:CO.pg.cutoff]

# visual
CO.pg.box <- boxplot(sorted.Ts[1:CO.pg.cutoff], main="Period Boxplot")

# the average influencial period
print('mean of top periods')
CO.pg.box.mean <- CO.pg.box$stats[3]
print(CO.pg.box.mean)
```

The average influencial period can be interpreted as on average the seasons have a period of 3.7 days

The top ten periods are about: 3.7, 2.3 and 5

The top ten frequencies are about: 0.27, 0.44, and 0.2
-this makes sense based on the periodogram
  - would like to add one more frequency at about 0.05
    - shows up in top 20 frequencies
    
### Assign Potential Periods to Variables
```{r}
# ideally these are not hard coded
CO.p1 <- 1/.27
CO.p2 <- 1/.44
CO.p3 <- 1/.2
CO.p4 <- 1/0.05

CO.p1
CO.p2
CO.p3
CO.p4
```

### Create Model with Potential Periods
```{r}
# create time variable
time.CO<-c(1:length(CO.ts))

# actual model
CO.lm.top4 <- lm(CO.ts ~ sin(2*pi*time.CO/CO.p1) + 
                         cos(2*pi*time.CO/CO.p1) +
                         sin(2*pi*time.CO/CO.p2) + 
                         cos(2*pi*time.CO/CO.p2) +
                         sin(2*pi*time.CO/CO.p3) + 
                         cos(2*pi*time.CO/CO.p3) +
                         sin(2*pi*time.CO/CO.p4) + 
                         cos(2*pi*time.CO/CO.p4))

# model summary
summary(CO.lm.top4)
```

### Compare Larger Model with Model with Only First Important Period
```{r}
# actual model
CO.lm.top1 <- lm(CO.ts ~ sin(2*pi*time.CO/CO.p1) + 
                         cos(2*pi*time.CO/CO.p1))

# model summary
summary(CO.lm.top1)

# anova
anova(CO.lm.top1, CO.lm.top4)
```

There is significant evidence to use the larger model

### Stepwise Regression on Larger Model
```{r}
CO.lm.step <- step(CO.lm.top4)
summary(CO.lm.step)
```

### Visual Inspection of Larger Model
```{r}
plot(CO.ts)
lines(CO.lm.top4$fitted.values, col = "red")

plot(CO.ts, xlim=c(0,100))
lines(CO.lm.top4$fitted.values, col = "red")

plot(CO.ts, xlim=c(100,200))
lines(CO.lm.top4$fitted.values, col = "red")
```

So the model is pretty bad, maybe we need to look at the first or second differences

### Visual Inspection of Step Model
```{r}
plot(CO.ts)
lines(CO.lm.step$fitted.values, col = "red")

plot(CO.ts, xlim=c(0,100))
lines(CO.lm.step$fitted.values, col = "red")

plot(CO.ts, xlim=c(100,200))
lines(CO.lm.step$fitted.values, col = "red")
```

step model also sucks

Conclusion
-Use the step function because it has the highest AIC

## Trend
### Model Time Series Based on Time
```{r}
# the actual model
CO.lm.trend <- lm(CO.ts ~ time.CO)

# summary analysis
summary(CO.lm.trend)
```

The p-value is 0.92 so there is not a significant trend in the data

## Autoregressive and Moving Average Components

## Model Assessment 
### Metrics
compare models based on adjusted R2 and/or AIC
### Diagnostics
diagnostics of linear models of trends + seasonality and ARIMA models of residuals
### Forecasting
compare models based on MSE

## Final Model

## Diagnostics
what problems remain in the diagnostics of the selected model?


# Univariate Time Series for NO2
## Visualize the Data
```{r}
# create time series
NO.ts <- ts(air.train$NO2.GT.)

# visualize raw data
plot(NO.ts)

# periodogram
pg.NO <- spec.pgram(NO.ts,spans=9,demean=T,log='no')

# acf and pacf
acf(NO.ts)
pacf(NO.ts)
```

No obvious trend or seasonality based on raw data

based on periodogram there may be some seasonlity, but it is going to be based on complex waves


# Multivariate Time Series
## Seasonality
## Trend
## Autoregressive and Moving Average Components
### Metrics
compare models based on adjusted R2 and/or AIC
### Diagnostics
diagnostics of linear models of trends + seasonality and ARIMA models of residuals
### Forecasting
compare models based on MSE

## Final Model

## Diagnostics
what problems remain in the diagnostics of the selected model?


# Simulation from Univariate CO Model
```{r}
# CO univariate model = model of data + model of residuals
CO.uni <- CO.final + CO.resi.final # idk that this is the way to do this

# simulate 1 year of daily maximum CO concentrations
### fix this based on actual ar and ma terms
CO.uni.sim <- arima.sim(n = 365*1, list(ar = c(CO.uni$coef[1], CO.uni$coef[2]), ma = c(CO.uni$coef[3])), sd = sqrt(CO.uni$sigma2))

# make time series
CO.uni.sim.ts <- ts(CO.uni.sim)
```

## Visualization
```{r}
# plot simulated values with observations
plot(CO.ts, main = "Simulation from Univariate CO Model", col = "black")
lines(CO.uni.sim.ts, col = "red")
legend(0.6, 0.57, legend = c("Time Series Data", "Simulation"), col = c("black", "red"))
```
how does simulation compare visually with time series data?

## Trend
```{r}
# linear model for simulation
CO.sim.lm <- lm()
```
compare coefficient estimates

## Seasonality
```{r}
# compare periodogram of observations and periodogram of simulation
par(mfrow = c(1,2))
pg.CO
pg.CO.uni.sim <- spec.pgram(CO.uni.sim.ts, spans = 9, demean = T, log = 'no')
par(mfrow = c(1,1))
```
compare periodograms (get periods?)

## Mean and Variance
```{r}
# mean of observations
mean(CO.ts)

# mean of simulation
mean(CO.uni.sim.ts)

# variance of observations
var(CO.ts)

# variance of simulation
var(CO.uni.sim.ts)
```
compare mean and variance

## Auto-Correlation
```{r}
# ACF of observations
acf(CO.ts)

# ACF of simulation
acf(CO.uni.sim.ts)
```
compare ACF

```{r}
# PACF of observations
pacf(CO.ts)

# PACF of simulation
pacf(CO.uni.sim.ts)
```
compare PACF

## Cross-Correlation
```{r}
# cross-correlation of observations
cor(air.train$CO.GT.)

# cross-correlation of simulation
cor(CO.uni.sim)
```
compare cross-correlation

# Simulation from Univariate NO2 Model
## Visualization
## Trend
## Seasonality
## Mean and Variance
## Auto-Correlation
## Cross-Correlation


# Simulation from Multivariate Model
## Visualization
## Trend
## Seasonality
## Mean and Variance
## Auto-Correlation
## Cross-Correlation

